Hierarchical Region-Network Sparsity for\\ High-Dimensional Inference in Brain Imaging

### 2016-02-05 paper/icml2016/example_paper.tex
gaels comments

\begin{abstract} 
Structured sparsity penalization has recently improved
statistical models applied to high-dimensional data in various domains.
%
As an extension to imaging neuroscience,
the present work incorporates
priors on network hierarchies of brain regions into
logistic-regression to distinguish neural activity.
This recombines neurobiological architectures underlying
functional segregation into brain regions and
functional integration by brain networks
that are studied separately in neuroscientific research.
%
Hierarchical region-network priors are shown to better classify
18 psychological tasks from a reference dataset
than other sparse estimators.
Weighing the relative importance of
region and network structure within the hierarchical tree penalty
recovered complementary aspects of the neural activity patterns.
%
Local and global priors of neurobiological knowledge
are thus demonstrated to enhance
generalization performance, sample complexity, and domain interpretability.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical ``long data" regime to
the high-dimensional ``wide data" regime.
% the high-dimensional regime (i.e., ``wide data") \cite{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This $n \ll p$ scenario challenges various statistical methods from
classical statistics.
For instance,
estimating generalized linear models without additional assumptions
yields an underdetermined system of equations.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{hastie2015statistical}.
They act as regularizer by encouraging zero coefficients
in model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Moreover, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus shrinking and selecting variables guided by
expected data distributions
\cite{bach2012optimization}.
Such restrictions to statistical complexity
are an attractive plan of attack
for the $\textgreater$100,000 variables per brain map.
Yet, what generally accepted neurobiological structure lends itself
to exploitation using structured sparsity priors?

@@

Functional specialization has long been
explored and interpreted based on different research methods.
%
For instance,
single-cell recordings and microscopic tissue examination
revealed the
segregation of the occipital visual cortex into V1, V2, V3, V3A/B, and V4
regions
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.
% INTEGRAL
It is more recent
that the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015}.
%
Besides analyses of
electrophysiological oscillations
% \cite{buzsaki2004neuronal}
and
graph-theoretical properties,
% graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity correlation \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
%
As a common point of this second set of methods,
interpretation of findings naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in contrast to methods examining regional specialization.

@@

Major brain networks emerge in human individuals around birth
\cite{doria2010}, before mental capacities mature in childhood. 
In adults,
the regions that compose
a given cohesive network are known to have more
similar functional profiles than nodes from different networks
\cite{anderson2013}.
As a well-established method for network estimation,
ICA \cite{beckmann2005} yields continuous brain maps with
weights for each voxel. The region nodes of ICA
networks are spatially disjoint sets of voxel groups that
agree with boundaries of brain atlases.
Such brain atlases can be obtained by segregating
the brain's grey matter into local voxel groups
using clustering algorithisms
\cite{cbp2015review}.
Consequently,
each region from a brain atlas can be uniquely associated with one of
the extracted ICA networks.
%
In the present work previously published network definitions
obtained using ICA \cite{smith2009}
and
previously published region definitions obtained from
spatially constrained ward clustering \cite{crad12}
allow constructing a hierarchy of global ICA networks with their
assigned local cluster regions
(Figure \ref{fig_priors}).
The ensuing network-region tree is used as a structural prior
of expected weight distributions
during supervised model fitting.

@@

% \paragraph{Problem formulation.}
\subsection{Problem formulation}
We formulate our estimation problem in the framework of regularized
empirical risk
minimization applied to linear models.
The goal is to estimate a good predictor of psychological task
given single brain images. Let the set \(\mathcal X\in\R^{nxp}\) represent brain
images of \(p > 0\) voxels.
%
We then minimize the risk \(\mathcal L(\hat y, y)\) with
\(\hat y = f(X\hat w + \hat b)\), where \(f\) is a link function 
(e.g., sigmoid for logistic regression, identity for linear regression),
and \(\mathcal L\) usually represents an appropriate negative loglikelihood.
We incorporate an informative prior through regularization:
\[\argmin_{w, b} \mathcal L(f(Xw + b), y) + \lambda\Omega(w),\]
where \(\lambda > 0\) and \(\Omega\) is the regularizer.

@@

\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$)
was quantitatively compared (Figure \ref{Tab:fig_dataratio}).
Region-network priors should contrain model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).

>>>>

\begin{abstract} 
Structured sparsity penalization has recently improved
statistical models applied to high-dimensional data in various domains.
%
As an extension to imaging neuroscience,
the present work incorporates
priors on network hierarchies of brain regions into
logistic-regression to distinguish neural activity.
These priors bridge two levels of neurobiological architecture 
that are studied separately in neuroscientific research:
functional segregation into brain regions and
functional integration by brain networks
%
Hierarchical region-network priors are shown to better classify
18 psychological tasks from a reference dataset
than other sparse estimators.
Varying the relative importance of
region and network structure within the hierarchical tree penalty
recovered complementary aspects of the neural activity patterns.
%
Local and global priors of neurobiological knowledge
are thus demonstrated to enhance
generalization performance, sample complexity, and domain interpretability.

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent shift from the classical ``long data" regime to
the high-dimensional ``wide data" regime.
% the high-dimensional regime (i.e., ``wide data") \cite{jordan2015massive}.
In the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations per data sample.
This $n \ll p$ scenario challenges various statistical methods from
classical statistics.
For instance,
estimating generalized linear models without additional assumptions
yields an underdetermined system of equations.
%
Many such ill-posed estimation problems
have benefited from
\textit{sparsity} assumptions
\cite{hastie2015statistical}.
They act as regularizer by encouraging zero coefficients
in model selection.
Sparse supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Moreover, \textit{structured sparsity} can impose
domain knowledge on the 
statistical estimation,
thus shrinking and selecting variables guided by
expected data distributions
\cite{bach2012optimization}.
Such restrictions to model complexity
are an attractive plan of attack
for the $\textgreater$100,000 variables per brain map.
Yet, what generally accepted neurobiological structure lends itself
to exploitation using structured sparsity priors?

@@

Functional specialization has long been
explored and interpreted based on different research methods.
%
For instance,
single-cell recordings and microscopic tissue examination
revealed the
segregation of the occipital visual cortex into V1, V2, V3, V3A/B, and V4
regions
\cite{hubel1962receptive, zeki1978functional}.
Tissue lesion of the mid-fusiform gyrus of the visual system,
in turn,
was frequently reported to impair
recognition of others' identity from faces
\cite{iaria2008contrib}.
%
As a crucial common point,
all these methodological approaches
yield neuroscientific findings
that are naturally interpreted according to
non-overlapping, discrete region compartments
as the basic architecture of brain organization.
% INTEGRAL
More recently,
the main interpretational focus has shifted
from circumscribed regions to network stratifications
in systems neuroscience \cite{yuste2015}.
%
Besides analyses of
electrophysiological oscillations
% \cite{buzsaki2004neuronal}
and
graph-theoretical properties,
% graph-theoretical properties \cite{bullmore2009complex},
studies of
functional connectivity correlation \cite{buckner2013opportunities} and
independent component analysis (ICA) \cite{beckmann2005}
became the workhorses of network discovery
in neuroimaging.
%
As a common point of this second set of methods,
interpretation of findings naturally embraces
cross-regional integration by
overlapping network compartments
as the basic architecture of brain organization,
in contrast to methods examining regional specialization.

@@

Major brain networks emerge in human individuals around birth
\cite{doria2010}, before mental capacities mature in childhood. 
In adults,
the regions that compose
a given cohesive network are known to have more
similar functional profiles than nodes from different networks
\cite{anderson2013}.
As a well-established method for network estimation,
ICA \cite{beckmann2005} yields continuous brain maps with
weights for each voxel. The region nodes of ICA
networks are spatially disjoint sets of voxel groups that
agree with boundaries of brain atlases.
Such brain atlases can be obtained by segregating
the brain's grey matter into local voxel groups
using clustering algorithisms
\cite{cbp2015review}.
Consequently,
each region from a brain atlas can be uniquely associated with one of
the extracted ICA networks.
%
In the present work previously published network definitions
obtained using ICA \cite{smith2009}
and
previously published region definitions obtained from
spatially constrained spectral clustering \cite{crad12}
allow constructing a hierarchy of global ICA networks with their
assigned local cluster regions
(Figure \ref{fig_priors}).
The ensuing network-region tree is used as a structural prior
of expected weight distributions
during supervised model fitting.

@@

% \paragraph{Problem formulation.}
\subsection{Problem formulation}
We formulate our estimation problem in the framework of regularized
empirical risk
minimization applied to linear models.
The goal is to estimate a good predictor of psychological task
given single brain images. Let the set \(\mathcal X\in\R^{n \times p}\) represent brain
images of \(p > 0\) voxels.
%
We then minimize the risk \(\mathcal L(\hat y, y)\) with
\(\hat y = f(X\hat w + \hat b)\), where \(f\) is a link function 
(e.g., sigmoid for logistic regression, identity for linear regression),
and \(\mathcal L\) usually represents an appropriate negative loglikelihood.
We incorporate an informative prior through regularization:
\[\argmin_{w, b} \mathcal L(f(Xw + b), y) + \lambda\Omega(w),\]
where \(\lambda > 0\) and \(\Omega\) is the regularizer.

@@

\subsection{Sample complexity of naive versus informed sparse model selection}
% \paragraph{Sample complexity of naive versus informed sparse model selection.}
Subsequently, the sample complexity of
$\ell_1$-penalized and hierarchical-tree-penalized logistic regression
($\alpha = 1, \beta = 1$)
was empirically estimated and quantitatively compared
(Figure \ref{Tab:fig_dataratio}).
Region-network priors should contrain model selection towards more
neurobiologically plausible classification estimators.
This should yield better out-of-sample generalization and
support recovery than
neurobiology-naive $\ell_1$-constrained logistic regression
in the data-scarce and data-rich scenarios.
%
The HCP data with examples from 18
psychological tasks were first divided into
90\% of training set (i.e., 7584 neural activity maps) and
10\% of test set (i.e., 842 maps).
Both learning algorithms were fitted based on the
training set at different subsampling fractions:
20\% (1516 neural activity maps),
40\% (3033 maps),
60\% (4550 maps),
80\% (6067 maps), and
100\% (7584 maps).
